---
title: "Trends and Patterns Associated with Survey Data"
author: \centering{Elliot Vosburgh\textsuperscript{1}, Samuel Costa\textsuperscript{1}, Nick Barone\textsuperscript{2} \ \vspace{5 mm} \textsuperscript{1} Department of Biology, University of Rhode Island \ \textsuperscript{2} Department of Kinesiology, University of Rhode Island}
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
fontsize: 11pt
gemoetry: margin=lin
keywords: data analysis, trends, datacarpentry
endnote: no
fig_caption: yes
thanks: Analysis performed on Ernest et al. 2017, whose original files are located
  at (https://github.com/weecology/portal-teachingdb).
abstract: This document is intended to identify data trends and patterns associated
  with the dataset of Ernest et. al 2017. All data analysis was performed using the
  R programming language and some select packages.
---

# Weight: Differences between Species and Sex over Time

An analysis of inter- and intra-species weights over time can provide details about how animals respond to environmental changes; although no environmental descriptions were given in the dataset, such as food abundancy, assumptions can be made based on the recorded observations.

In Figure 1, mean weight for each recorded species is plotted over individual years. Some species show relatively low weight changes, while others display highly varied weights. The authors propose that these animals, NL, DS, and SH, inhabit more specific or more threatened niches than the other survey species.

```{r, fig1, warning=FALSE, echo = FALSE, message=FALSE, fig.width = 7, fig.asp = .6, fig.align = "center", fig.pos = 'h', fig.cap = "Interspecific mean weight over time."}

# Load necessary libraries
library(tidyverse)
library(directlabels)

# Import data from CSV file
surveys_complete <- read_csv("data_output/surveys_complete.csv")

# Store average weights of species_id by year

yearly_weight <- surveys_complete %>%
  group_by(year, species_id) %>%
  summarize(avg_weight = mean(weight))

# Render a weight-comparison plot with sexes combined

ggplot(data = yearly_weight, aes(x = year, y = avg_weight, color = species_id)) +
  geom_line(size = 1.5, alpha = 0.5) +
  geom_dl(aes(label = species_id), method = list(dl.trans(x = x + 0.1), "last.points"), cex = 1.0) +
  theme_grey(base_family = "") +
  theme(axis.text.x = element_text(size = 11, angle = 0),
        axis.text.y = element_text(size = 11, angle = 0),
        text = element_text(size = 11),
        legend.position = "none") +
  labs(x = "Year",
       y = "Weight")

```

The weights for these three species do not seem to be linked, so it may be supposed that they inhabit different niches, affected differently by habitat changes. Huge swings are recorded for SH, with weight dipping to almost 50 units from over 100 units in under 5 years. Similarly, weight for NL starts at over 175 in 1976, but falls to 140 in 1978. The only correlation seems to be that their average weights begin to rise around 1997 - after that, NL begins to gain while SH begins to lose. No data for DS after this rise was available after 1998.

```{r, fig2, warning=FALSE, echo=FALSE, message=FALSE, fig.width = 7, fig.asp = .6, fig.align = "center", fig.pos = 'h', fig.cap = "Intraspecific mean weight over time, separated by sex."}

# Load necessary libraries
library(tidyverse)
library(directlabels)

# Import data from CSV file
surveys_complete <- read_csv("data_output/surveys_complete.csv")

# Store average weights of sexes of species_id by year

yearly_sex_weight <- surveys_complete %>%
  group_by(year, sex, species_id) %>%
  summarize(avg_weight = mean(weight))

# Render a weight-comparison plot with sexes split

ggplot(data = yearly_sex_weight, aes(x = year, y = avg_weight, color = species_id)) +
  geom_line(size = 1.5, alpha = 0.5) +
  facet_grid(sex ~ .) +
  geom_dl(aes(label = species_id), method = list(dl.trans(x = x + 0.1), "last.points"), cex = 0.1) +
  theme_grey(base_family = "") +
  theme(axis.text.x = element_text(size = 11, angle = 0),
        axis.text.y = element_text(size = 11, angle = 0),
        text = element_text(size = 11),
        legend.position = "none") +
  labs(x = "Year",
       y = "Weight")

```

So there are interspecific differences in mean weight over time. What of intraspecific variance? Males and females for many terrestrial species vary greatly in weight, but what does the data say about intraspecies resilience?

In Figure 2, mean weight was separated out by sex. The authors found that the same species which exhibited interspecific variance, NL, DS, and SH, also exhibited major swings in average weight between males and females. In NL, males were more vulnurable to weight changes, and were faster to gain or lose weight than females. The same can be said for DS and SH. Whether this vulnurability is physiological or behavioral, that is, males willingly forgoing food and giving it to females, is conjecture without further information.
